\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[ruled]{algorithm2e}
\usepackage{hyperref}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\newcommand{\pimattn}{PIM-Attn}
\newcommand{\kivi}{KiVI}

\begin{document}

\title{\pimattn{}: Efficient Asymmetric Quantized Attention with Near-Bank Processing-in-Memory}

\author{\IEEEauthorblockN{Anonymous Author(s)}
\IEEEauthorblockA{\textit{Anonymous Institution}\\
Anonymous City, Anonymous Country\\
anon@example.com}
}

\maketitle

\begin{abstract}
Large Language Models (LLMs) face a critical memory bottleneck from their key-value (KV) cache during inference.
While recent works like \kivi{} propose tuning-free asymmetric 2-bit quantization to dramatically reduce KV cache size, the dequantization overhead on GPU introduces significant latency.
Existing near-bank processing-in-memory (PIM) approaches are insufficient for the mixed-precision operations required by asymmetric quantization.
We propose \pimattn{}, a heterogeneous attention framework that decomposes the asymmetric dot product into parallel 2-bit operations mapped to near-bank PIM units, while routing float16-intensive score-value operations to GPU.
Cycle-accurate simulation on Llama-3-8B demonstrates $1.4\times$--$2.9\times$ per-sequence Q@K latency improvement over GPU-only execution, with $2.1\mu s$ per-sequence latency on 2-bit KV using 2048 PIM banks.
Our results show that direct 2-bit PIM compute is $2\times$ more efficient than dequantize-then-compute approaches, validating the decomposition strategy.
\end{abstract}

\begin{IEEEkeywords}
LLM inference, KV cache quantization, near-bank PIM, attention, heterogeneous computing
\end{IEEEkeywords}

\input{intro}
\input{background}
\input{method}
\input{experiment}
\input{related}

\section{Conclusion}
We presented \pimattn{}, a heterogeneous attention framework that maps decomposed asymmetric quantized attention operations to near-bank PIM. Our key insight is that the asymmetric dot product $Q\cdot K \approx s_Q s_K[\sum \hat{q}\hat{k} - z_K\sum\hat{q} - z_Q\sum\hat{k} + d z_Q z_K]$ decomposes into 2-bit-compatible operations ideal for near-bank PIM, while float16-intensive score@V operations remain on GPU. Cycle-accurate simulation on Llama-3-8B demonstrates $1.4\times$ per-sequence Q@K speedup over GPU. Future work includes extending to prefill-mode attention and integrating with tiling-based memory optimization.

\bibliographystyle{IEEEtran}
\bibliography{ref}

\end{document}
