%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is the template for submission
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[hyphens]{url}
\usepackage{multirow}
\usepackage{adjustbox}
\usepackage{xspace}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xfrac}
\usepackage{hhline}
\usepackage{balance}
\usepackage[all]{nowidow}
%\usepackage{hyperref}
%\usepackage{doi}
\usepackage[bookmarks=true,breaklinks=true,letterpaper=true,colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}


\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}


% Ensure letter paper
\pdfpagewidth=8.5in
\pdfpageheight=11in


\pagenumbering{arabic}


%%%%%% -- PAPER CONTENT STARTS-- %%%%%%%%


% --- Acronyms ---
\newcommand{\tbt}{{TBT}\xspace}
\newcommand{\ttft}{{T2FT}\xspace}
\newcommand{\ete}{{E2E}\xspace}

\newcommand{\genstage}{{decode-only stage}\xspace}
\newcommand{\deconlystage}{{decoding-only stage}\xspace}
\newcommand{\mixedstage}{{mixed stage}\xspace}

\newcommand{\duplex}{{Duplex}\xspace}
\newcommand{\bankbundle}{{bank bundle}\xspace}
\newcommand{\highp}{{xPU}\xspace}
\newcommand{\lowp}{{Logic-PIM}\xspace}
\newcommand{\bankpim}{{Bank-PIM}\xspace}
\newcommand{\bgpim}{{Bankgroup-PIM}\xspace}

\newcommand{\opb}{{Op/B}\xspace}

\newcommand{\hgxb}{{\textbf{HGX}}\xspace}
\newcommand{\gpu}{{GPU}\xspace}
\newcommand{\doublegpu}{{2$\times$GPU}\xspace}

\newcommand{\duplexh}{{Duplex-Hetero}\xspace}
\newcommand{\duplexpe}{{Duplex+PE}\xspace}
\newcommand{\duplexpet}{{Duplex+PE+ET}\xspace}
\newcommand{\duplexplus}{{Duplex+Plus}\xspace}
\newcommand{\duplexbankpim}{{Bank-PIM}\xspace}
\newcommand{\duplexsplit}{{Duplex-Split}\xspace}


\newcommand{\mixtral}{{Mixtral}\xspace}
\newcommand{\dsmoe}{{DeepSpeed-MoE}\xspace}
\newcommand{\glam}{{GLaM}\xspace}
\newcommand{\grok}{{Grok1}\xspace}
\newcommand{\opt}{{OPT}\xspace}
\newcommand{\llama}{{Llama3}\xspace}

\newcommand{\ie}{\textit{i.e.}\xspace}
\newcommand{\eg}{\textit{e.g.}\xspace}
\newcommand{\cf}{\textit{cf.}\xspace}
\newcommand{\Numex}{$N_{ex}$\xspace}
\newcommand{\deggrp}{$deg_{grp}$\xspace}
\newcommand{\topk}{top-$k$\xspace}
\newcommand{\lin}{$L_{in}$\xspace}
\newcommand{\lout}{$L_{out}$\xspace}
\newcommand{\nhead}{$N_{head}$\xspace}

% --- graph & table ---
\newcolumntype{M}[1]{>{\centering\arraybackslash}m{#1}}



%%%%%%%%%%%---SETME-----%%%%%%%%%%%%%
\title{Duplex: A Device for Large Language Models with Mixture of Experts, Grouped Query Attention, and Continuous Batching} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%%%%%%%%%% -- User Defined Start -- %%%%%%%%%%%%%
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{xspace}
\usepackage{hhline}
\usepackage{makecell}
\usepackage{multirow}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{lipsum}
\usepackage{blindtext}
\usepackage{tikz}
%\usepackage{xfrac}
%\usepackage{booktabs}
\usepackage{subfigure}
\usepackage{enumitem}
%%%
\usepackage{color}
\usepackage{soul}
\usepackage{comment}
\usepackage{balance}
\usepackage{xfrac}
\usepackage{dblfloatfix}
\usepackage{mathabx}
\usepackage{amsmath}
\usepackage{dashrule}

\usepackage[all]{nowidow}

\newcommand{\sungmin}[1]{{\color{magenta} #1}}
\newcommand{\jongmin}[1]{{\color{blue} #1}}
\newcommand{\kwanhee}[1]{{\color{orange} #1}}
\newcommand{\jwchoi}[1]{{\color{magenta} #1}}
%%%%%%%%%%% -- User Defined End -- %%%%%%%%%%%%%


\author{\IEEEauthorblockN{Sungmin Yun\IEEEauthorrefmark{2},
Kwanhee Kyung\IEEEauthorrefmark{2},
Juhwan Cho\IEEEauthorrefmark{2}, 
Jaewan Choi\IEEEauthorrefmark{2},
Jongmin Kim\IEEEauthorrefmark{2},\\
Byeongho Kim\IEEEauthorrefmark{3},
Sukhan Lee\IEEEauthorrefmark{3},
Kyomin Sohn\IEEEauthorrefmark{3},
Jung Ho Ahn\IEEEauthorrefmark{2}}
\IEEEauthorblockA{\IEEEauthorrefmark{2}\textit{Seoul National University, Seoul, South Korea}, \IEEEauthorrefmark{3}\textit{Samsung Electronics, Hwasung, South Korea}\\
\textit{\{sungmin.yun, kwanhee5, jfcho2, cjw9202, jongmin.kim, gajh\}@snu.ac.kr}\\ \textit{\{bh1122.kim, sh1026.lee, kyomin.sohn\}@samsung.com}}
}


\begin{document}
\maketitle
\pagestyle{plain}


%%%%%% -- PAPER CONTENT STARTS-- %%%%%%%%

\begin{abstract}
Large language models (LLMs) have emerged due to their capability to generate high-quality content across diverse contexts.
To reduce their explosively increasing demands for computing resources, a mixture of experts (MoE) has emerged.
The MoE layer enables exploiting a huge number of parameters with less computation.
Applying state-of-the-art continuous batching increases throughput; however, it leads to frequent DRAM access in the MoE and attention layers.
We observe that conventional computing devices have limitations when processing the MoE and attention layers, which dominate the total execution time and exhibit low arithmetic intensity (\opb).
Processing MoE layers only with devices targeting low-\opb such as processing-in-memory (PIM) architectures is challenging due to the fluctuating \opb in the MoE layer caused by continuous batching.



To address these challenges, we propose \duplex, which comprises \highp tailored for high-\opb and \lowp to effectively perform low-\opb operation within a single device. 
\duplex selects the most suitable processor based on the \opb of each layer within LLMs.
As the \opb of the MoE layer is at least 1 and that of the attention layer has a value of 4--8 for grouped query attention, prior PIM architectures are not efficient, which place processing units inside DRAM dies and only target extremely low-\opb (under one) operations.
Based on recent trends, \lowp adds more through-silicon vias (TSVs) to enable high-bandwidth communication between the DRAM die and the logic die and place powerful processing units on the logic die, which is best suited for handling low-\opb operations ranging from few to a few dozens.
To maximally utilize the \highp and \lowp, we propose expert and attention co-processing.
By exploiting proper processing units for MoE and attention layers, \duplex shows up to 2.67$\times$ higher throughput and consumes 42.0\% less energy compared to GPU systems for LLM inference.

\end{abstract}

\begin{IEEEkeywords}
large language models, processing-in-memory, mixture of experts, grouped query attention, continuous batching
\end{IEEEkeywords}

\input{1_introduction}
\input{2_background}
\input{3_motivation}
\input{4_contribution-1}
\input{5_contribution-2}
\input{6_experimental_setup}
\input{7_evaluation}
\input{8_discussion}
\input{9_related_work}
\input{10_conclusion}

\section*{Acknowledgment}
This work was partly supported by the Samsung Electronics, the National Research Foundation of Korea grant funded by the Korea government (MSIT) (RS-2024-00405857), Institute of Information \& communications Technology Planning \& Evaluation (IITP) grant funded by MSIT (RS-2024-00402898, RS-2021-II211343), and IITP under the artificial intelligence semiconductor support program funded by MSIT (IITP-2023-RS-2023-00256081).
The EDA tool was supported by the IC Design Education Center (IDEC), Korea.
The ICT at Seoul National University provides research facilities for this study.
Jung Ho Ahn is the corresponding author.

%%%%%%% -- PAPER CONTENT ENDS -- %%%%%%%%


%%%%%%%%% -- BIB STYLE AND FILE -- %%%%%%%%
\balance
\bibliographystyle{IEEEtranS}
\bibliography{refs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

